---
title: "6. Miscellaneous"
author: "Edzer Pebesma"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{6. Miscellaneous}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```
This vignette describes a number of issues that did not come up in
the previous vignettes, and that may or may not be categorized as
"frequently asked questions". Readers are encouraged to provide
entries for this vignette (as for the others).

# What is this EPSG code all about?

EPSG stands for a maintained, well-understood registry of spatial
reference systems, maintained by the International Association
of Oil \& Gass Producers (IOGP). "EPSG" stands for the authority,
e.g. "EPSG:4326" stands for spatial reference system with ID 4326
as it is maintained by the EPSG authority. The website for the EPSG
registry is found at the epsg.org domain.

# How does `sf` deal with secondary geometry columns?

`sf` objects can have more than one geometry list-column,
but always only one geometry column is considered _active_,
and returned by `st_geometry`.  When there are multiple
geometry columns, the default `print` methods reports which
one is active:
```{r}
library(sf)
demo(nc, ask = FALSE, echo = FALSE)
nc$geom2 = st_centroid(st_geometry(nc))
print(nc, n = 2)
```

We can switch the active geometry by using `st_geometry<-` or `st_set_geometry`, as in 
```{r}
plot(st_geometry(nc))
st_geometry(nc) <- "geom2"
plot(st_geometry(nc))
```

# Does `st_simplify` preserve topology?

`st_simplify` is a topology-preserving function, but does this on the
level of individual feature geometries. That means, simply said, that
after applying it, a polygon will still be a polygon. However when
two features have a longer shared boundary, applying `st_simplify`
to the object does not guarantee that in the resulting object these
two polygons still have the same boundary in common, since the
simplification is done independently, _for each feature geometry_.

# Why do dplyr verbs not work for `sf` objects?

They do! However, many developers like to write scripts that never
load packages but address all functions by the `sf::` prefix, as in
```{r,eval=FALSE}
i = sf::st_intersects(sf1, sf2)
```

This works up to the moment that a `dplyr` generic like `select` for an `sf` object
is needed: should one call `dplyr::select` (won't know it should search
in package `sf`) or `sf::select` (which doesn't exist)? Neither works.
One should in this case simply load `sf`, e.g. by
```{r,eval=FALSE}
library(sf)
```

# What is this error/warning/message about?

## _although coordinates are longitude/latitude, xxx assumes that they are planar_

Most (but not all) of the geometry calculating routines used by `sf` come from the [GEOS](https://libgeos.org) library. This library considers coordinates in a two-dimensional, flat, Euclidean space. For longitude latitude data, this is not the case. A simple example is a polygon enclosing the North pole, which should include the pole:
```{r}
polygon = st_sfc(st_polygon(list(rbind(c(0,80), c(120,80), c(240,80), c(0,80)))), 
		crs = 4326)
pole = st_sfc(st_point(c(0,90)), crs = 4326)
st_intersects(polygon, pole)
```

which gives a wrong result (no intersection).

## _st\_centroid does not give correct centroids for longitude/latitude data_

Similar to the above, centroids are computed assuming flat, 2D space:
```{r}
st_centroid(polygon)[[1]]
```

where the centroid should have been the pole.

## _dist is assumed to be in decimal degrees (arc\_degrees)._

This message indicates that `sf` assumes a distance value is given in degrees. To avoid this message, pass a value with the right units:
```{r}
pt = st_sfc(st_point(c(0,0)), crs = 4326)
buf = st_buffer(polygon, 1)
buf = st_buffer(polygon, units::set_units(1, degree))
```

