---
title: "7. Spherical geometry in sf using s2geometry"
author: "Edzer Pebesma and Dewey Dunnington"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{7. The s2geometry library for spherical geometry in sf}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```
This vignette describes what spherical geometry implies, and how
package `sf` uses the s2geometry library (https://s2geometry.io)
for geometrical measures, predicates and transformations.

Spatial coordinates either refer to _projected_ (or Cartesian)
coordinates, meaning that they are associated to points on a flat
space, or to unprojected or _geographic_ coordinates, when they
refer to angles (latitude, longitude) pointing to locations on a
sphere (or ellipsoid). The flat space is also referred to as R2,
the sphere as S2.

Package `sf` implements _simple features_, a standard for point,
line, and polygon geometries where geometries are built from points
(nodes) connected by straight lines (edges). The simple feature
standard does not say much about its suitability for dealing with
geographic coordinates, but the topological relational system it
builds upon ([DE9-IM](https://en.wikipedia.org/wiki/DE-9IM)) refer
to R2, the two-dimensional flat space. 

Yet, more and more data are routinely served or exchanged using
geographic coordinates. Using software that assumes an R2, flat
space may work for some problems, and although `sf` up to version 0.9-x
had some functions in place for spherical/ellipsoidal computations
(from package `lwgeom`, for computing area,
length, distance, and for segmentizing), it has also happily warned 
the user that it is doing R2, flat computations with such coordinates with messages like
```
although coordinates are longitude/latitude, st_intersects assumes that they are planar
```
hinting to the responsibility of the user to take care of potential
problems. Doing this however leaves ambiguities, e.g. whether 
`LINESTRING(-179 0,179 0)`

* passes through `POINT(0 0)`, or
* passes through `POINT(180 0)`

and whether it is
* a straight line, cutting through the Earth's surface, or
* a curved line following the Earth's surface

Starting with `sf` version 1.0, `sf` uses the new package `s2`
(Dunnington, Pebesma, Rubak 2020) for spherical geometry, which
has functions for computing pretty much all measures, predicates
and transformations _on the sphere_. This means:

* no more hodge-podge of some functions working on R2, with annoying messages, some on the ellipsoid
* a considerable speed increase for some functions
* no computations on the ellipsoid (which are considered more accurate, but are also slower)

The `s2` package is really a wrapper around the C++
[s2geometry](https://s2geometry.io) library which was written by
Google, and which is used in many of its products (e.g. Google
Maps, Google Earth Engine, Bigquery GIS) and has been translated
in several programming other languages.

# Fundamental differences

Compared to geometry on R2, and DE9-IM, the `s2` package brings a
few fundamentally new concepts, which are discussed first.

## Polygons on S2 divide the sphere in two parts

On the sphere (S2), any polygon defines two areas; when following the
exterior ring, we need to define what is inside, and the definition
is _the left side of the enclosing edges_. This also means that
we can flip a polygon (by inverting the edge order) to obtain the
other part of the globe, and that in addition to an empty polygon
(the empty set) we can have the full polygon (the entire globe).

Simple feature geometries should obey a ring direction too: exterior
rings should be counter clockwise, interior (hole) rings should
be clockwise, but in some sense this is obsolete as the difference
between exterior ring and interior rings is defined by their position
(exterior, followed by zero or more interior). `sf::read_sf` has an
argument `check_ring_dir` that checks, and corrects, ring directions
and many (legacy) datasets have wrong ring directions. With wrong
ring directions, many things still work.

For S2, ring direction is essential. For that reason, `st_as_s2`
has an argument `oriented = FALSE`, which will check and correct
ring directions, assuming that all exterior rings occupy an area
smaller than half the globe:
```{r}
library(sf)
nc = read_sf(system.file("gpkg/nc.gpkg", package="sf")) # wrong ring directions
library(s2)
s2_area(st_as_s2(nc, oriented = FALSE)[1:3]) # corrects ring direction, correct area:
s2_area(st_as_s2(nc, oriented = TRUE)[1:3]) # wrong direction: Earth's surface minus area
nc = read_sf(system.file("gpkg/nc.gpkg", package="sf"), check_ring_dir = TRUE)
s2_area(st_as_s2(nc, oriented = TRUE)[1:3]) # no second correction needed here:
```

## Half-closed polygon boundaries 

Polygons in `s2geometry` can be
* CLOSED: they contain their boundaries, and a point on the boundary intersects with the polygon
* OPEN: they do not contain their boundaries, points on the boundary do not intersect with the polygon
* HALF-CLOSED: they contain part of their boundaries, but no boundary of non-overlapping polygons is contained by more than one polygon.

In principle the DE9-IM model deals with interior, boundary and
exterior, and intersection predicates are sensitive to this (the
difference between _contains_ and _covers_ is all about boundaries).
DE9-IM however cannot uniquely assign points to polygons when
polygons form a polygon _coverage_ (no overlaps, but mostly common
boundaries). This means that if we would count points by polygon,
and some points fall _on_ shared polygon boundaries, we either miss
them (_contains_) or we count them double (_covers_); this leads to
bias or need for post-processing. Using HALF-CLOSED non-overlapping
polygons guarantees that every point is assigned to _maximally_
one polygon in an intersection. This corresponds to e.g. how this
would be handled in a grid (raster) coverage, where every grid cell
(typically) only contains its upper-left corner and its upper and
left sides.

```{r}
a = as_s2_geography("POINT(0 0)")
b = as_s2_geography("POLYGON((0 0,1 0,1 1,0 1,0 0))")
s2_intersects(a, b, s2_options(model = 0)) # OPEN
s2_intersects(a, b, s2_options(model = 2)) # CLOSED
s2_intersects(a, b, s2_options(model = 1)) # HALF-CLOSED: a toss
s2_intersects(a, b) # default: HALF-CLOSED
```

## Cap, enclosing rectangle

Computing the minimum and maximum values over coordinate ranges,
as `sf` does with `st_bbox()`, is of limited value for spherical
coordinates because

* small regions covering the antimeridian end up with a huge longitude range
* regions including a pole will end up with a latitude range not extending to +/- 90

S2 has two alternatives: the cap and the `lat_lng_rect`:

```{r}
fiji = s2_data_countries("Fiji")
aa = s2_data_countries("Antarctica")
s2_cap(fiji)
s2_lat_lng_rect(c(fiji,aa))
```

Note that for Fiji, `lng_lo` being higher than `lng_hi` indicates that
the region covers (crosses) the antimeridian.

# Switching between S2 and GEOS

The two-dimensional R2 library that was formerly used by `sf` is
[GEOS](https://trac.osgeo.org/geos/), and `sf` can be instrumented to
use GEOS or `sf`. First we will ask if `s2` is being used by default:
```{r}
sf_use_s2()
```
then we can switch it of (and use GEOS) by
```{r}
sf_use_s2(FALSE)
```
and switch it on (and use S2) by
```{r}
sf_use_s2(TRUE)
```



# Measures

## Area
```{r}
library(sf)
library(units)
nc = read_sf(system.file("gpkg/nc.gpkg", package="sf"))
sf_use_s2(TRUE)
a1 = st_area(nc)
sf_use_s2(FALSE)
a2 = st_area(nc)
plot(a1, a2)
abline(0, 1)
summary((a1 - a2)/a1)
```

## Length
```{r}
nc_ls = st_cast(nc, "MULTILINESTRING")
l1 = st_length(nc_ls)
l2 = st_length(nc_ls, use_lwgeom = TRUE)
plot(l1 , l2)
abline(0, 1)
summary((l1-l2)/l1)
```


## Distances
```{r}
d1 = st_distance(nc, nc[1:10,])
d2 = st_distance(nc, nc[1:10,], use_lwgeom = TRUE)
dim(d1)
dim(d2)
plot(as.vector(d1), as.vector(d2))
abline(0, 1)
summary(as.vector(d1)-as.vector(d2))
```

# Predicates

# Transformations

## References

* Dewey Dunnington, Edzer Pebesma and Ege Rubak, 2020. s2:
Spherical Geometry Operators Using the S2 Geometry Library.
https://r-spatial.github.io/s2/, https://github.com/r-spatial/s2

